---
title: "Economic Time Series Analysis"
author: "David Alexander Moe"
date: "2023"
output: github_document
---

The analysis concerns the industrial production index (base 2015 = 100), which measures the change over time in the physical volume (output) of production carried out by industry. The monthly value is derived from a special sample survey conducted among Italian industrial enterprises. The time series runs from January 1990 to October 2021 (382 observations). The data can be found on the 'Eurostat' website: <https://ec.europa.eu/eurostat/data/database>.

The objective of this analysis is to forecast the change in the industrial production index for the next 12 months (short horizon), from November 2021 to October 2022.

For obvious reasons, many things were omitted leaving only the relevant parts.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Importing necessary dependencies

Importing libraries and functions that will be useful throughout the analysis

```{r}
rm(list = ls())

# libraries and functions
library(forecast)
library(tseries)
library(tsoutliers)
library(urca)
library(FinTS)
```

Look at the "Functions" folder to see all the specific details about the functions.

```{r}
source("G:/Il mio Drive/TRIENNALE/TS Analysis/Economic-TSA-ARIMA/Functions/TSA-Useful-Functions.R")
source("G:/Il mio Drive/TRIENNALE/TS Analysis/Economic-TSA-ARIMA/Functions/CalendarEffects-Functions.R")
source("G:/Il mio Drive/TRIENNALE/TS Analysis/Economic-TSA-ARIMA/Functions/TSA-Predict-Functions.R")
```

## Read data and declare some external variables

```{r}
filedati <- "G:\\Il mio Drive\\TRIENNALE\\TS Analysis\\Economic-TSA-ARIMA\\Production in industry - monthly data (Italy).csv"

#### Read data
dati <- read.table(file = filedati, header = TRUE, sep = ",", quote = "", 
                   na.strings = ".", check.names = FALSE, comment.char = "")
name <- "Industrial production in Italy"

dati$TIME_PERIOD <- paste(dati$TIME_PERIOD,"-15",sep="")
dati$TIME_PERIOD <- gsub("-","",dati$TIME_PERIOD)
dati$TIME_PERIOD <- as.Date(x=dati$TIME_PERIOD, format= "%Y %m %d")

start <- as.numeric( c( format(dati$TIME_PERIOD[1], "%Y"), format(dati$TIME_PERIOD[1], "%m") ) )

y <- ts(data=dati$OBS_VALUE, start= start, frequency=12)
```

```{r}
# some external variables

#### calendar effects
cal <- .calendarEffects(time = dati$TIME_PERIOD, country = "it")

#cal <- cal[, c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat","Sun","sh", "lh", "eh"), drop = FALSE]
cal <- cal[, c("wd", "lh", "eh"), drop = FALSE]

cal <- as.matrix(cal)

#### Drift if needed
drift <- cbind(drift = 1 : NROW(y))
```

## Preliminary analysis

```{r}
#### Ts plot, acf, pacf of the original series
# par(mfrow = c(3,1))
x <- y
plot(x, type = "l", main = name, ylab = "Index Number",xlab="Time")
```

The graph depicts the time series: there is a strong annual seasonality and an upward trend until 2008, the year of the economic crisis. Following this, the industrial production index maintained an almost constant level until the Covid-19 period.

```{r}
Acf(x = x, type = "correlation", na.action = na.pass, lag.max = 60, main = name)
Acf(x = x, type = "partial",     na.action = na.pass, lag.max = 60, main = name)
```

Given the monthly seasonality, a seasonal ARIMA model (p, d, q)x(P, D, Q)[S] with period S=12 is used for the analysis. These graphs show the ACF and PACF respectively, i.e. the autocorrelation and partial autocorrelation at different lags which are indicated in years. The ACF graph confirms the presence of seasonality, given the decay of ACF at multiple lags of S, while the PACF graph denotes that $\psi(12)$ is close to 1: it would appear that the stochastic process is non-stationary due to the seasonal part (D\>0). To be sure of this non-stationarity, the Augmented Dickey-Fuller (ADF) test is performed.

```{r}
#### UNIT ROOT TESTS:
#### DF/ADF tests with trend
#### The following rows show how to make DF test in practice.
##   DF is however not recommended because gives severely biased results 
##   in case the DGP is more complex than the equation estimated by DF.
##   This time series prove this (compare DF with ADF results)
df.1 <- ur.df(y = y, type = "trend", lags = 0, selectlags = "Fixed") 
df.2 <- ur.df(y = y, type = "drift", lags = 0, selectlags = "Fixed") 
df.3 <- ur.df(y = y, type = "none",  lags = 0, selectlags = "Fixed")


#### (DGP:   RW + drift; 
##    Model: AR(1) + trend (+ other possible stationary terms))

adf.1 <- ur.df(y = y, type = "trend", lags = 24, selectlags = "AIC")
print( adf.1@teststat )
print( adf.1@cval )
##   1) H0(1) -> tau3 A and H0(2) -> phi3 A => no trend, go to 2   
##   2) H0(3) -> phi2 A                     => no drift, go to 3   

#### (DGP:   RW; 
##    Model: AR(1) + drift (+ other possible stationary terms))

adf.2 <- ur.df(y = y, type = "drift", lags = 24, selectlags = "AIC")
print( adf.2@teststat )
print( adf.2@cval )
```

Accepting all tested H0s, $$H0(1): \gamma=0, H0(2): \xi=\gamma=0, H0(3): \omega=\xi=\gamma=0, H0(5): \gamma=0 , H0(6): \omega=\gamma=0$$ (the last 2 H0s obtained by switching to DGP=RW and $\xi=0$ in the model) we conclude that at least one unit root is definitely present. From the data, it appears that the unit root found is due to the seasonal part.

Therefore, a new ADF test on twelfths differences is carried out to identify other possible URs:

```{r}
dod <- diff(y,lag=12)
adf.3 <- ur.df(y = dod, type = "trend", lags = 24, selectlags = "AIC")
print( adf.3@teststat )
print( adf.3@cval )
```

From which we note that there are not other URs.

## ARIMA estimation

Based on the preliminary analyses, the starting point for modelling ARIMA is the setting of d=0 and D=1, after which p, q, P and Q are formulated. From here, after various considerations based on Information Criteria, residual analysis, likelihood values and coefficient estimation, it is concluded that the best model without external regressors is:

```{r}
#### ARIMA (no external regressors)
xreg <- NULL
fit <- Arima(y = y, 
             order = c(3, 0, 0), seasonal = list(order = c(2, 1, 2)),
             xreg = xreg, include.constant = FALSE)
print(summary(fit))
fit1 <- fit
```

By inserting external regressors, in particular calendar variables and possible anomalies, the selected model is:

```{r}
#### ARIMA + external regressors (var di calendario)
xreg <- cal
fit <- Arima(y = y,
             order = c(3, 0, 0), seasonal = list(order = c(0, 1, 1)),
             xreg = xreg, include.constant = FALSE)
print(summary(fit))
fit2 <- fit
```

The inclusion of the variable 'working days' (wd) is motivated by the fact that only Sundays and single holidays were not significant in the model with all calendar variables.

```{r}
#### Root analysis
#parte ar(p) con A(L)=0 e ma(q) con B(L)=0
par(mfrow = c(1,2))

#fit1
root <- .arma.roots(fit = fit1)
.circle(win = 2)
points(root$root$ar, col = "red")
points(root$root$ma, col = "blue")
legend("bottomright", inset= 0, title="ARIMA", 
       legend=c("AR roots", "MA roots"),
       col=c("red", "blue"), pch=1, cex=.70)

#fit2
root <- .arma.roots(fit = fit2)
.circle(win = 2)
points(root$root$ar, col = "red")
points(root$root$ma, col = "blue")
legend("bottomright", inset= 0, title="ARIMA + Cal", 
       legend=c("AR roots", "MA roots"),
       col=c("red", "blue"), pch=1, cex=.70)

mtext("Root Analysis Comparison", side = 3, line = -1.5, outer = TRUE)


### fit1 shows MA roots (stemming from the seasonal part) critically close to 1
### in module.
### invertible MA part (roots B(L)B*(L^s)=0) -> here arima is invertible
### some ar roots very close to those ma -> lower Q from 2 to 1 in the model+cal and try again
### much better results by clearly having more separated roots
```

Using the powerful tso() function in the models specified so far yields the anomalies. In these chunks of code is presented a portion of the ARIMA modeling with some external regressors.

```{r}
################################################################################
## ARIMA modeling with anomalies
################################################################################

#####################################
## ARIMA
#####################################

#### Copy model
fit <- fit1
#### Extract settings
settings <- .Arima.settings(fit = fit)
#### Prepare xreg
xreg <- NULL

#### Fit
fit <- tso(y = y, xreg = xreg,
           types = c("AO", "LS", "TC"), delta = 0.7, cval = 5,
           maxit = 10, maxit.iloop = 100, maxit.oloop = 10,
           tsmethod = "arima",
           args.tsmethod = list( order = settings$order, seasonal = settings$seasonal) )
fit1.o <- fit

#### Reporting
print(fit)
plot(fit)

#### Extract outlier effects fro  the above model
oeff <- outliers.effects(mo = fit$outliers, n = NROW(y), pars = coef(fit$fit),
                         weights = FALSE)

## Plot weighted effects
par(mfrow=c(1,1))
plot(x = dati$TIME_PERIOD, main="Anomalies weighted effects", xlab="Year", ylab="effect", y = rowSums(oeff), type = "l")

#### Estimate again
xreg <- as.matrix( oeff )
fit <- Arima(y = y,
             order = settings$order, seasonal = settings$seasonal,
             include.constant = settings$include.constant,
             xreg = xreg)
fit3 <- fit

```

```{r}
#####################################
## ARIMA + calendar effects
#####################################

#### Copy model
fit <- fit2
#### Extract settings
settings <- .Arima.settings(fit = fit)
#### Prepare xreg
xreg <- cal
#xreg <- if ( settings$include.drift ) { cbind(drift, xreg) } else { xreg }
#### Fit
fit <- tso(y = y, xreg = xreg,
           types = c("AO", "LS", "TC"), delta = 0.7, cval = 5,
           maxit = 10, maxit.iloop = 100, maxit.oloop = 10,
           # tsmethod = "auto.arima",
           # args.tsmethod = list(allowdrift = false, ic = "bic", trace = true) )
           tsmethod = "arima",
           args.tsmethod = list( order = settings$order, seasonal = settings$seasonal) )
fit2.o <- fit

### Reporting
print(fit)
plot(fit)
### Extract elements
## Outlier effects
oeff <- outliers.effects(mo = fit$outliers, n = NROW(y), pars = coef(fit$fit),
                         weights = FALSE)
## Plot weighted effects
par(mfrow=c(1,1))
plot(x = dati$TIME_PERIOD, main="Anomalies weighted effects", xlab="Year", ylab="effect", y = rowSums(oeff), type = "l")
```

The outliers graph shows the presence of 3 outliers: an LS from December 2008, and a TC and AO at the start of the Covid-19 pandemic (March-April 2020). The anomalies have a substantial impact since the ICs of the simple ARIMA integrated with anomalies are much lower than the model with calendar variables. As a result of these considerations, it can be conjectured that the best model is the ARIMA integrated with calendar variables and anomalies.

```{r}
#### Estimate again
xreg <- cbind(cal, oeff)
fit <- Arima(y = y,
             order = settings$order, seasonal = settings$seasonal,
             include.constant = settings$include.constant,
             xreg = xreg)
fit4 <- fit

#### Root analysis dei modelli con anomalie
#parte ar(p) con A(L)=0 e ma(q) con B(L)=0
par(mfrow = c(1,2))

#fit3
root <- .arma.roots(fit = fit3)
.circle(win = 2)
points(root$root$ar, col = "red")
points(root$root$ma, col = "blue")
legend("bottomright", inset= 0, title="ARIMA + Outliers", 
       legend=c("AR roots", "MA roots"),
       col=c("red", "blue"), pch=1, cex=.70)

#fit4
root <- .arma.roots(fit = fit4)
.circle(win = 2)
points(root$root$ar, col = "red")
points(root$root$ma, col = "blue")
legend("bottomright", inset= 0, title="ARIMA + Calendar + Outliers", 
       legend=c("AR roots", "MA roots"),
       col=c("red", "blue"), pch=1, cex=.70)

mtext("Root Analysis Comparison", side = 3, line = -1.5, outer = TRUE)

```

As expected, the model that best fits the data is the last one, which has the same anomalies as the model without calendar variables. This last graph shows the roots of the AR and MA characteristic equations. In the first model, they are very close to each other: reshaping the ARMA orders solves this problem.

## Diagnostics

```{r}
fit <- fit4

#### Useful quantities
npar1  <- NROW(fit$coef)                            ## Number of parameters
lag1   <- npar1 + c(1, 2, 5, 10, 15, 20)
res1   <- residuals(fit)                            ## Residuals
resst1 <- ( res1 - mean(res1) ) / sqrt(fit$sigma2)  ## Standardized residuals

#### Ts plot, acf, pacf, Ljung-Box of residuals
par(mfrow = c(2,1), cex.main=1.15)
main <- "Residuals"
x1 <- res1
plot(x1, type = "l", main = main, ylab = "Residuals")
Acf(x = x1, type = "correlation", lag.max = 60, na.action = na.pass, main = main)


# lb <- Box.test(x = res1, lag = 10, type = "Ljung-Box", fitdf = NROW(fit$coef))
cat("\n", paste("Ljung-Box of", main, "on different lags\n") )
lb <- mapply(FUN = Box.test, lag = lag1, 
             MoreArgs = list(x = x1, type = "Ljung-Box", fitdf = npar1))[1:3, , drop = FALSE]
print(rbind(lag1,lb))
### The test is highly significant at all lags, although the ACF does not show much room for improvement 



#### Ts plot, acf of residuals^2
main <- "Residuals^2"
x1 <- res1^2
plot(x1, type = "l", main = main, ylab = "")
Acf(x = x1, type = "correlation", lag.max = 60, na.action = na.pass, main = main)

#### Ts plot, acf of |residuals|
main <- "|Residuals|"
x1 <- abs(res1)
plot(x1, type = "l", main = main, ylab = "")
Acf(x = x1, type = "correlation", lag.max = 60, na.action = na.pass, main = main)

```

These Graphs show the plot and ACF of the model residuals integrated with calendar variables and anomalies. They show that they have a behaviour compatible with a white noise. The ACF of the squared residuals and \|residuals\|, leads to the assumption that they are generated by independent random variables. The autocorrelations almost all fall within the bands.

Let's check the ARCH test on\|residuals\|:

```{r}
#### Another diagnostic: the ARCH test
cat("\n-----------------------------------------------------------------
  ARCH based preliminary analyses\n")
cat("ARCH test on demeaned log-returns\n")
lag <- c(1, 2, 3, 6, 12, 24)
at <- mapply(FUN = ArchTest, lags = lag, 
             MoreArgs = list(x = x1 , demean = TRUE))
print(at[1:3,])
```

Rejecting the null hypothesis at almost all M chosen past values leads to the conclusion that the conditional variance is not entirely stable, indicating that perhaps a transformation of the data is necessary to stabilise it.

Here, the transformation test which suggests not transforming the data:

```{r}
trt <- .trsf.test(fit = fit4, msg = "Transformation check on 'fit4'\n")
```

```{r}
#### unconditional distribution of the standardised residuals
## Plot
par(mfrow = c(1,2))
hist(x = resst1, breaks = 25, freq = FALSE, main = "Residuals", xlab = "")
x1 <- seq(from = min(resst1), to = max(resst1)+1, length.out = 100) 
lines(x = x1, y = dnorm(x = x1, mean = 0, sd = 1), col = "red")
qqnorm(y = resst1, main = "Normal Q-Q Plot",
       xlab = "Theoretical quantiles", ylab = "Sample quantiles",
       plot.it = TRUE)
abline(a = 0, b = 1, col = "red")

## Test di normalità
print( shapiro.test(x = res1 ) )
```

Here the unconditional distribution of the standardised residuals: it denotes a certain leptokurticity and thus the Gaussian distribution does not fit the data well. Indeed, in the Shapiro-Wilk normality test, the null hypothesis of normality of the residuals is rejected (W = 0.97978, p-value = 3.464e-05).

## Forecast

### Ex-post forecasts: all made 1-step ahead

```{r}
#### Settings
J  <- 12                                              ## How many ex-post forecast to compute
H  <- 1                                               ## Forecasting horizon
t1 <- .predict.t1(nobs = NROW(y), J = J, n.ahead = H) ## 1st obs in the ex-post period (needed below)


#### No external regressors
pred1.1 <- .predict(object = fit1, n.ahead = H, t = t1, y = y,
                    fixed.n.ahead = TRUE)


#### If we have external regressors
newxreg <- cal
pred2.1 <- .predict(object = fit2, n.ahead = H, t = t1, y = y, xreg = newxreg,
                    fixed.n.ahead = TRUE)

#### If we have outliers
newxreg <- .oeff.4.predict(object = fit1.o, n.ahead = 0)
pred3.1 <- .predict(object = fit3, n.ahead = H, t = t1, y = y, xreg = newxreg,
                    fixed.n.ahead = TRUE)

#### If we have external regressors and outliers
x2 <- .oeff.4.predict(object = fit2.o, n.ahead = 0)
newxreg <- as.matrix(cbind(cal, x2))
pred4.1 <- .predict(object = fit4, n.ahead = H, t = t1, y = y, xreg = newxreg,
                    fixed.n.ahead = TRUE)

#### Naive
predn.1 <- .predict.naive(fit = fit4, J = J, n.ahead = H)

#### Bands
x1 <- .pred.bands(pred = pred1.1, alpha = 0.05)
x2 <- .pred.bands(pred = pred2.1, alpha = 0.05)
x3 <- .pred.bands(pred = pred3.1, alpha = 0.05)
x4 <- .pred.bands(pred = pred4.1, alpha = 0.05)

#### Error Measures
em1.1  <- .ErrorMeasures(y = y, fit = x1$mean, naive = predn.1)
em2.1  <- .ErrorMeasures(y = y, fit = x2$mean, naive = predn.1)
em3.1  <- .ErrorMeasures(y = y, fit = x3$mean, naive = predn.1)
em4.1  <- .ErrorMeasures(y = y, fit = x4$mean, naive = predn.1)
emn.1  <- .ErrorMeasures(y = y, fit = predn.1, naive = predn.1)
## Print
ErrorMeas <- data.frame(
  model = c("Arima", "Arima + Calendar", "Arima + Outliers", "Arima + Calendar + Outliers", "Naive"),
  h = H,
  rbind( em1.1, em2.1, em3.1, em4.1, emn.1, deparse.level = 0 ) )
print( ErrorMeas )
```

The naive in the table is a benchmark model that we take as a reference for comparisons: in the case of the data considered, the naive behaves like a seasonal RW because the process has seasonal integration. The model that has the greatest gain in forecasting is ARIMA with calendar effects and outliers because, when comparing all specified models with naive, it has the lowest scaled measurement errors.

This next Graph shows the ex-post forecasts. It turns out that the integrated model with calendar variables and outliers is the best because the true observations are all within the prediction bands of this model.

```{r}
#### Plot
ind  <- (NROW(y) - J + 1) : NROW(y)
ind1 <- 1 : NROW(ind)
par(mfrow = c(1,1))
ylim <- range(
  x1$lower[ind1], x1$upper[ind1], x2$lower[ind1], x2$upper[ind1],
  x3$lower[ind1], x3$upper[ind1], x4$lower[ind1], x4$upper[ind1] )
time <- dati$TIME_PERIOD[ind]
plot(x = time, y = y[ind], ylim = c(60,125),
     main = "(Ex-post) Past 12 months forecasts", xlab = "time", ylab = "Production")
lines(x = time, y = x1$mean[ind1],  col = "red")
lines(x = time, y = x2$mean[ind1],  col = "cyan")
lines(x = time, y = x3$mean[ind1],  col = "violet")
lines(x = time, y = x4$mean[ind1],  col = "blue")
lines(x = time, y = predn.1[ind1],  col = "green", lty = "solid")
lines(x = time, y = x4$lower[ind1], col = "blue", lty = "dotted")
lines(x = time, y = x4$upper[ind1], col = "blue", lty = "dotted")
legend("bottomleft", inset=0, title="Arima: ", 
       legend=c("Simple","+ cal", "+ anomalies", "+ cal + outliers", "naive"), 
       col=c("red","cyan", "violet", "blue" ,"green"), lty=1, cex=0.75)
```

### Ex-ante forecasts

```{r}
#######################################################
## Ex-ante (genuine) forecasts: from 1 to H steps ahead
#######################################################

#### Settings
H  <- 12      ## Maximum forecasting horizon
t1 <- NROW(y) ## Last obs in the info set (T)

#### Time generation
## Useful for extending calendar and for plotting
time1 <- .extend.time(x = dati$TIME_PERIOD, n.ahead = H, by = "month")

#### No external regressors
newxreg <- NULL
pred1 <- .predict(object = fit1, n.ahead = H, t = t1, y = y, xreg = newxreg,
                  fixed.n.ahead = FALSE)

#### with external regressors
x1 <- .calendarEffects(time = time1)[, colnames(cal), drop = FALSE]
#I make the external v ar that I need for the forecast
#time1 what is outside the forecast period
newxreg <- as.matrix(rbind(cal, x1))
pred2 <- .predict(object = fit2, n.ahead = H, t = t1, y = y, xreg = newxreg,
                  fixed.n.ahead = FALSE)

#### outliers
x2 <- .oeff.4.predict(object = fit1.o, n.ahead = H)
newxreg <- x2
pred3 <- .predict(object = fit3, n.ahead = H, t = t1, y = y, xreg = newxreg,
                  fixed.n.ahead = FALSE)

#### external regressors + outliers
x2 <- .oeff.4.predict(object = fit2.o, n.ahead = H)
newxreg <- as.matrix( cbind( rbind(cal, x1), x2) )
pred4 <- .predict(object = fit4, n.ahead = H, t = t1, y = y, xreg = newxreg,
                  fixed.n.ahead = FALSE)

#### Naive
predn <- .predict.naive(fit = fit4, J = H, n.ahead = H) 

#### Bande
x1 <- .pred.bands(pred = pred1, alpha = 0.05)
x2 <- .pred.bands(pred = pred2, alpha = 0.05)
x3 <- .pred.bands(pred = pred3, alpha = 0.05)
x4 <- .pred.bands(pred = pred4, alpha = 0.05)


print( cbind(t = x1$t, pred1 = x1$mean, pred2 = x2$mean, pred3 = x3$mean, pred4 = x4$mean, 
             naive = predn) )
```

```{r}
#### Plot
par(mfrow = c(1,1))
ylim <- range(
  x1$lower, x1$upper, x2$lower, x2$upper,
  x3$lower, x3$upper, x4$lower, x4$upper, predn )
time <- time1
plot(x = time, y = x1$mean, type = "l", col = "red",
     main = "(Ex-ante) Forecasts for the next 12 months", xlab = "time", ylab = "Production",
     ylim = ylim)
lines(x = time, y = x2$mean,  col = "cyan")
lines(x = time, y = x3$mean,  col = "violet")
lines(x = time, y = x4$mean,  col = "blue")
lines(x = time, y = x4$lower, col = "blue", lty = "dotted")
lines(x = time, y = x4$upper, col = "blue", lty = "dotted")
lines(x = time, y = predn, col = "green")
legend("bottomleft", inset=0, title="Arima: ", 
       legend=c("Simple","+ cal", "+ anomalies", "+ cal + outliers", "naive"), 
       col=c("red","cyan", "violet", "blue" ,"green"), lty=1, cex=0.75)
```

The ex-ante forecasts follow similar trends. In particular, a negative peak in production is observed around August 2022, which is likely to occur.
